---
title: "Archaeological Spatial Analysis in R"
subtitle: "Lesson 1"
author:
  - name: Michal Michalski
    email: michal.m.michalski@durham.ac.uk
    affiliations: Durham University
format: 
  html:
    code-fold: false
    toc: true
    code-overflow: wrap
editor: visual
execute: 
  warning: false
theme: cosmo
---

Learning Objectives

-   learn about R-spatial ecosystem

-   introduce vector and raster data

-   plot spatial data that are used in the class

## Introduction

The objective of this lesson is to read and plot spatial data in R.

You will also become familiar with syntax variety of packages used for mapping.

## Background Reading

-   Lovelace R, Nowosad K, Muenchow J, *Geocomputation with R,* Chapter 1,2 and 9 - [link](https://r.geocompx.org/)

## R-spatial

CRAN Task View: Analysis of Spatial Data provides a comprehensive list of most of R packages created to work with spatial data.

<https://cran.r-project.org/web/views/Spatial.html>

Fundamental are R packages with classes representing spatial data:

-   vector: `sf`, `sp`, `stars`, `spatstats`

-   raster: `terra`, `raster`, `stars`

## Data

The data were obtained from various publications and open source repositories.

Vector data for the classes are stored in a **geopackage** data.gpkg**.**

`data>vect`

Raster data are stored in **geotiff** format.

`data>rast`

Follow this link to download the data on your computer.

## Vector

The sf package

We are going to use package sf to read and map vector layers by using the following functions:

-   `st_layers()` returns layers in a data source

-   `st_read()` reads simple features from file or database.

-   `plot()` plots one or more attributes

-   `head()` return first parts of the object

Load sf package

```{r}
# install.packages("sf")
library(sf)
```

Explore content of the geopackage

```{r}

st_layers("../data/vect/data.gpkg")

```

**survey_extent** - 12 km radius covering area surveyed around Tell Beydar

**source:** Ur and Wilkinson 2008

```{r}
# read
survey_extent = st_read("../data/vect/data.gpkg", layer = "survey_extent", quiet = TRUE)

# glimpse
# head(survey_extent)

# plot
plot(survey_extent["code"], lty = "dashed", col = NA)
```

**sites_point** - a point layer of archaeological settlements surveyed around Tell Beydar

**source:** Ur and Wilkinson 2008; Palmisano et al. - [link](https://doi.org/10.5281/zenodo.4294558)

```{r}

# read
sites = st_read("../data/vect/data.gpkg", layer = "sites_point", quiet = TRUE)

# glimpse
# head(sites, n = 3)

#plot
plot(st_geometry(sites))

```

**sites_polygon -** a polygon layer of archaeological settlements detected using remote sensing based survey; the polygons were derived from sites outlines and locations.

**source:** Menze and Ur, 2012 - [link](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/7H8K3N)

```{r}

# read
sites_polygon = st_read("../data/vect/data.gpkg", layer = "sites_polygon", quiet = TRUE)

# glimpse
# head(sites_polygon, n = 3)

#plot
plot(sites_polygon[,"max_elevation_mound_m"], 
     nbreaks = 10, 
     pal = hcl.colors(11, "Cividis"),
     lwd = 0.1,
     key.pos = 4)
```

**hollow_ways** - ancient tracks now visible as shallow linear depressions; represented as polylines.

**source:** Ur, 2010 - [source](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/NYWQO2)

```{r}
# read
hollow_ways = st_read("../data/vect/data.gpkg", layer = "hollow_ways", quiet = TRUE)

# glimpse
# head(hollow_ways, n = 3)

# convert numeric to factor
hollow_ways$hw_type = as.factor(hollow_ways$hw_type)

#plot
plot(hollow_ways[,"hw_type"])
```

**rimes_rivers** - a watercourse derived from 'Rivers of Mesopotamia' dataset

**source:** Bätscher, 2020 - [link](https://doi.org/10.5281/zenodo.3787265)

```{r}

# read
rivers = st_read("../data/vect/data.gpkg", layer = "rimes_rivers", quiet = TRUE)

# glimpse
# head(rivers, n = 3)

#plot
plot(st_geometry(rivers), col = "#80c5de")
```

**geology**: polygon layer derived from geology dataset for the Jazira Region of Syria.

**source:** Mathys, 2017 - [link](https://doi.org/10.7488/ds/1795)

```{r}
#read
geology = st_read("../data/vect/data.gpkg", layer = "geology")

# glimpse
# head(geology, n = 3)

#plot
plot(geology[,"SEDIMENTS"], key.width = lcm(6), key.pos = 4, pal = hcl.colors(4, "Roma"))

```

## Raster

Raster data can be an are handled by [terra](https://cran.r-project.org/web/packages/terra/index.html) package. It provides classes and methods to read raster, execute raster algebra expression or regression type models.

We will now explore raster data used in this class using the following functions:

-   `rast()` creates a SpatRaster object

-   `plot()` base function to plot raster data

-   `plotRGB()` allows to map RGB color image

Load package

```{r}

# install.packages("terra")
library(terra)

```

Read digital elevation model (DEM) - single layer

```{r}

# read in dem
dem = rast("../data/rast/dem.tif")

# print dem to look at the summary
dem

```

Plot raster

```{r}

plot(dem , main = "Digital Elevation Model")
```

Read Landsat image - multi-layer

```{r}

landsat = rast("../data/rast/landsat_tbs/landsat.tif")

landsat
```

Plot function will display all 3 bands

```{r}

plot(landsat)

```

We can use `plotRGB()` to plot Red-Green-Blue image from 3 bands of Landsat image.

```{r}

plotRGB(landsat, r = "band_4", g = "band_3", b = "band_2", stretch = "lin")

```

## Mapping

#### Data

Load and tidy data for mapping

```{r}
# r packages
library(dplyr)
library(terra)
library(tmap)
library(sf)

# survey
survey_extent = st_read("../data/vect/data.gpkg", layer = "survey_extent", quiet = TRUE)

# rivers
rivers = st_read("../data/vect/data.gpkg", layer = "rimes_rivers", quiet = TRUE)

# hollow_ways
hollow_ways = st_read("../data/vect/data.gpkg", layer = "hollow_ways", quiet = TRUE)

# geology
geology = st_read("../data/vect/data.gpkg", layer = "geology", quiet = TRUE)

# select basalt sediment
basalt = subset(geology, SEDIMENTS == "Basalts", quiet = TRUE)

# sites 
sites = st_read("../data/vect/data.gpkg", layer = "sites_point", quiet = TRUE) 

# raster
dem = rast("../data/rast/dem.tif")

# color
c3t1 <- tidyterra::hypsometric_tints_db %>%
  filter(pal == "c3t1")

# hillshade
slope = terrain(dem, 'slope', unit='radians')

aspect = terrain(dem, 'aspect', unit='radians')

hillshade = shade(slope, aspect, 30, 315)


```

#### mapsf

basic plot

```{r}

#install.packages(mapsf)
library(mapsf)

# theme from arguments
mf_init(x = survey_extent)
# Plot the base map
mf_map(x = survey_extent)
# Plot proportional symbols
mf_map(x = sites, var = "size_ha", type = "prop")

```

advanced plot

```{r}

# set theme
mf_theme("default")
# Initiate a base map
mf_init(x = survey_extent)
# plot survey_extent 
mf_map(survey_extent, add = TRUE)
# plot population
mf_map(
  x = sites, 
  var = "size_ha",
  type = "grad",
  breaks = c(0,5,10,20,45),
  col = "brown4",
  leg_pos = "bottomleft2",  
  leg_title = "Area in Hectares"
)
# annotation
mf_annotation(
  x = c(641692.6, 4066955),
  txt = "Tell Beydar",
  pos = "topleft", cex = 1.2, font = 2,
  halo = TRUE, s = 1.5
)
# layout
mf_layout(title = "TBS Survey - Sites distribution", 
          credits = paste0("Sources: Ur and Wilkinson 2008\n",
                           "mapsf ", 
                           packageVersion("mapsf")))


```

#### ggplot

basic plot

```{r}

# install.packages("ggplot2")
library(ggplot2)

# plot
ggplot() + 
  geom_sf(data = survey_extent) +
  geom_sf(data = sites,aes(size =  size_ha)) +
  coord_sf(datum = st_crs(sites))

```

advanced plot

```{r}
# library to handle SpatRaster 
library(tidyterra)

# plot
ggplot() + 
  geom_spatraster_contour_filled(data = dem, show.legend = FALSE) +
  geom_sf(data = survey_extent, colour = "black",lty = "dashed", fill = NA) +
  geom_sf(data = sites,aes(size =  size_ha)) +
  scale_fill_whitebox_d(palette = "atlas") +
  scale_size(name = "Size Ha") +
  coord_sf(datum = st_crs(sites)) +
  labs(
    title = "Tell Beydar Survey",
    caption = "Source: Ur and Wilkonson 2008"
  ) +
  theme_void()
```

#### tmap

basic plot

```{r}

# install.packages("tmap")
library(tmap)

# simple map
tm_shape(survey_extent) +
  tm_borders() +
  tm_shape(sites) +
  tm_bubbles(size = "size_ha")


```

advanced plot

```{r}
# hillshade
(map = tm_shape(hillshade) +
  tm_raster(palette = gray(0:100 / 100), n = 100, legend.show = FALSE) +
# dem
  tm_shape(dem) +
  tm_raster(alpha = 0.7, palette = c3t1$hex,legend.show = FALSE, n = 10) +
# survey extent
  tm_shape(survey_extent) +
  tm_borders(lwd = 2,lty = "dashed", col = "#5f8171") +
# watercourse
  tm_shape(rivers) +
    tm_lines(col = "#80c5de", lwd = 3) +
# hollow ways
  tm_shape(hollow_ways) +
    tm_lines(col = "#5f8171", alpha = 0.2, lwd = 1) +
# geology 
  tm_shape(basalt) +
    tm_borders(lty = "dashed") +
# sites
  tm_shape(sites) +
  tm_bubbles(col = "size_ha",
             shape = 19, 
             size = "size_ha",
             palette = "black",
             legend.col.show = FALSE,
             title.size = "Site Size (Ha)",
             legend.size.is.portrait = TRUE,
             scale = 1.5,
             legend.hist = T
             ) +
# layout
  tm_scale_bar(breaks = c(0,2,4), position = c("left", "bottom")) +
  tm_credits("Source:Ur and Wilkinson 2008",) +
  tm_layout(
    main.title = "Tell Beydar Archaeological Survey",
    frame = TRUE, 
    inner.margins = 0,
    legend.outside = TRUE,
    legend.hist.width = 1,
    legend.hist.height = 0.8))

# save map
# tmap_save(map, "assets/map.png")

```

## Exercise

1.  Go back to your tmap basic map and develop the code further.
2.  Consult the code in advanced example abd Chapter 9 of *Geocomputation with R.*
